---
title: "MESSI: Multi Experiments with SyStematic Interrogation"
# Remove the old title
# "A Nextflow pipeline for benchmarking multiomics integration methods for disease classification"
author:
  - "Chunqing (Tony) Liang"
  - "Tajveer Grewal"
  - "Asees Singh"
  - "Amrit Singh"
csl: assets/nature.csl
#csl: assets/springer-vancouver-brackets.csl
bibliography: assets/reference.bib
linkcolor: blue
link-citations: true
geometry: margin=2cm # Margin should be at least 2cm
nocite: '@*'
output: 
  bookdown::pdf_document2:
    number_sections: false
    toc: false
    extra_dependencies: ["flafter"]
    keep_md: true
    keep_tex: true
    includes:
      in_header: common_header.tex
header-includes:
  - \usepackage{lscape}
  - \usepackage{pdfpages}
  - \usepackage{graphicx}
  - \usepackage[figuresright]{rotating}
  - \usepackage{caption}
  # Use smaller font size and sans serif for figure captions
  - \captionsetup[figure]{font={small,sf,it}}

---

```{r setup, echo=FALSE, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(here)
library(cowplot)
library(kableExtra)
# Set global options for code chunks

# https://stackoverflow.com/questions/74538080/cannot-change-size-of-images-using-include-graphics-in-r-markdown

# You need to change the size of displayed figures by setting the chunk options out.width/out.height. 
# You can set only one of these two options if you want to keep the aspect ratio of the figures.
knitr::opts_chunk$set(echo=FALSE, fig.align = "center", results="asis", eval=F)

# Convenient variables
fig_path <- here("results/figures")
```

<!--- Add custom styling for figure captions  -->
<style>
p.caption {
  font-size: 0.1em;
}
</style>

# Abstract

Technological advances enable multiomics profiling of molecular data (e.g., genes, proteins, metabolites) from biological samples at bulk, single-cell, spatial resolution. Integrative methods identify shared patterns and biomarkers, improving disease understanding and clinical strategies. However, method selection is challenging due to varying analytical tasks (e.g., clustering, prediction) and data types. Existing reviews are task-specific or non-reproducible. We propose an automated framework in Nextflow, *MESSI*, for systematic benchmarking of multiomics methods. It ensures reproducibility, standardizes inputs across tasks, and runs on any computing platform, enabling reliable method evaluation and selection.

**Keywords**: bioinformatics pipeline, multiomics integration methods, machine learning, benchmark, multimodal learning



# Background

```{r background-child, child=c('sections/background.Rmd'), eval=TRUE}
```


# Results



```{r results-child, child=c('sections/results.Rmd'), eval=TRUE}
```

# Discussion

MESSI represents a paradigm shift in how the multiomics community approaches method evaluation and development. Unlike traditional benchmarking studies that provide static snapshots of method performance, MESSI establishes a living infrastructure that can continuously incorporate new methods as they emerge. This is crucial in a rapidly evolving field where novel machine learning and deep learning approaches are constantly being proposed. The modular, containerized architecture means that any researcher developing a new integration method can immediately benchmark against the established landscape without rebuilding evaluation pipelines from scratch. This dramatically lowers the barrier to rigorous method comparison and will accelerate the pace of methodological innovation by providing immediate feedback on whether new approaches represent genuine advances over existing methods.

MESSI directly addresses a critical challenge in computational biology: the lack of standardized, reproducible benchmarking frameworks. Our containerized approach ensures that method comparisons remain valid across different computing environments and time points, addressing a major limitation of previous reviews that relied on custom scripts with limited portability. The standardized MAE/MuData format we propose could serve as a community standard, reducing the technical barriers that currently prevent researchers from exploring multiple integration approaches. This standardization is particularly important given the rapid pace of method development in multiomics, where new approaches frequently emerge but are difficult to compare against existing methods due to implementation and data format differences.


MESSI's greatest impact will be its role in fostering community-wide methodological progress. By providing a common evaluation platform, it enables cumulative knowledge building that has been difficult to achieve with disparate, non-comparable benchmarking studies. Method developers can focus their efforts on genuine algorithmic innovations rather than reimplementing evaluation infrastructure. The framework also democratizes access to comprehensive benchmarking—smaller research groups without extensive computational expertise can now rigorously evaluate their methods against established baselines. This leveling of the playing field should accelerate innovation by enabling more researchers to contribute meaningful methodological advances. As the collection of evaluated methods grows, MESSI will become an increasingly valuable resource for understanding the methodological landscape and identifying promising directions for future development.

While our current evaluation focused on five established methods, the specific findings serve primarily to demonstrate MESSI's capabilities rather than provide definitive method rankings. Our comprehensive benchmarking reveals several critical insights into multiomics integration method performance that illustrate the framework's analytical power. The consistent superiority of DIABLO across datasets can be attributed to its supervised framework that explicitly leverages phenotype information during dimension reduction, contrasting with unsupervised approaches like MOFA that must rely on subsequent supervised learning steps. The design matrix optimization in DIABLO allows for explicit modeling of inter-omics relationships, which proves particularly valuable when signal strength is moderate, a common scenario in real-world datasets. Conversely, the weaker performance of deep learning approaches like MOGONET, despite their theoretical capacity for capturing complex non-linear relationships, highlights the challenges of training sophisticated models on relatively small sample sizes typical in multiomics studies.

The framework's ability to capture both predictive performance and computational resource requirements provides the multidimensional evaluation necessary for informed method selection. The substantial computational differences observed between methods have important practical implications. While Cooperative Learning and MOGONET demand significantly higher computational resources, this cost is not consistently justified by improved performance, particularly for datasets with moderate sample sizes. The efficient performance of DIABLO and MOFA suggests that for routine analyses, these methods offer an optimal balance of predictive power and computational feasibility. This is particularly relevant for clinical applications where rapid turnaround times are essential, or for exploratory studies with limited computational budgets. The memory usage patterns we observed also inform deployment strategies where methods with high prediction-time memory requirements may be unsuitable for real-time clinical decision support systems.

Our analysis reveals that method selection should be informed by specific dataset characteristics rather than relying on universal recommendations. The simulation studies demonstrate that when signal strength is low (signal = 3), most methods achieve reasonable performance (AUC $\sim 0.7$), but as datasets approach the "no signal" regime common in exploratory studies, method choice becomes critical. Dataset size emerges as a key determinant: cooperative learning methods show more stable performance across varying sample sizes, while complex methods like MOGONET require larger datasets to achieve their potential. Feature dimensionality ratios (samples to features) also influence method suitability, with regularized approaches like DIABLO maintaining robustness even in high-dimensional settings. These findings provide practitioners with concrete guidance for method selection based on their specific data characteristics.


Our findings reveal several opportunities for future methodological development that MESSI is uniquely positioned to evaluate. The observation that no single method dominates across all scenarios suggests that ensemble approaches, combining the strengths of different methods, represent a promising research direction. The consistent performance of regularized approaches indicates that sparsity-inducing methods are well-suited to multiomics data characteristics. Additionally, the moderate performance of even the best methods on real datasets with low signal strength highlights the need for methods specifically designed for exploratory analysis of subtle biological signals. Future users adding new methods whether it is novel deep learning architectures, ensemble approaches, or domain-specific adaptations will benefit from this same comprehensive evaluation capability, creating an ever-expanding knowledge base of method performance characteristics.


# Conclusion


We proposed MESSI, *Multi Experiment SyStematic Interrogation*,  an automated framework that  establishes the foundational infrastructure necessary for systematic, reproducible method evaluation in multiomics or multimodal integration. The framework's containerized, modular architecture creates a sustainable platform that will remain valuable as methods evolve and datasets grow more complex. By standardizing data formats, evaluation protocols, and computational environments, MESSI eliminates the technical barriers that have historically prevented comprehensive method comparisons, those standardized MAE/MuData formats leads to method exploration across the research community. While our current evaluation of five methods provides valuable insights—highlighting DIABLO's consistent performance and revealing computational trade-offs among approaches—the specific findings are less important than demonstrating MESSI's capability to generate such insights systematically. The framework enables researchers to make informed, data-driven decisions about method selection based on their specific dataset characteristics and computational constraints. More importantly, it provides method developers with immediate feedback about their innovations' performance relative to existing approaches, accelerating the cycle of methodological improvement and ensuring that new methods represent genuine advances over existing baselines.

The modular architecture of MESSI facilitates several important extensions that will keep pace with the rapidly evolving field: adaptation to single-cell and spatial multiomics data, incorporation of emerging deep learning architectures, and development of ensemble methods that combine complementary approaches. The continuous benchmarking capability positions MESSI to track method performance as datasets grow larger and more complex, providing ongoing guidance to the research community. This extensibility ensures that MESSI will function as a living repository of methodological knowledge rather than a static comparison that quickly becomes obsolete.

Most importantly, MESSI establishes a methodological standard for rigorous evaluation in multiomics integration, potentially serving as a model for benchmarking frameworks in other areas of computational biology where method proliferation has outpaced systematic evaluation. By providing this foundational infrastructure, MESSI enables the multiomics community to build cumulative knowledge about method performance, ultimately advancing the field from a collection of disparate approaches toward a mature analytical discipline with established best practices and rigorous evaluation standards. This transformation from ad hoc method comparisons to systematic, reproducible benchmarking represents a critical step toward evidence-based method selection and development in multiomics research.


\newpage

<!----

Here simply include child doc of method

---->

# Methods

```{r method-child, child=c('sections/methods.Rmd'), eval=TRUE}
```

\newpage

# References

::: {#refs}
:::

# (APPENDIX) Appendix {.unnumbered}

<!-- # More information -->

<!-- This will be Appendix A. -->

<!-- This proposed pipeline ensures full reproducibility of any method by using an independent container environment with Singularity [@kurtzer2017singularity]. We propose to publish this pipeline on the nf-core [@ewels2020nf] platform, a community of Nextflow users, to ensure easy access and usability for researchers worldwide. 


Currently, we are benchmarking existing methods like DIABLO [@singh2019diablo], Cooperative Learning [@ding2022cooperative], MOGONET [@wang2021mogonet], RGCCA [@girka2023multiblock], MOFA [@Argelaguet2018; @Argelaguet2020] . The current pipeline focuses on testing this selection of integrative methods against a series of real multiomics datasets, particularly for classification tasks.

The results are compared using metrics like balanced classification error, F1 score, and biological enrichment results.

Regarding classification performance, we employed the area under the curve (AUC) score for method comparison across datasets (Fig 1B). Mogonet emerged as the weakest performer, with MOFA + glmnet showing only marginal improvement. In contrast, Cooperative Learning demonstrated the highest performance, followed by DIABLO and RGCCA. Interestingly, DIABLO with a null design outperformed than its full design choice. Nevertheless, all methods performed poorly on the tcga-thca dataset.




Through MESSI, we could systematically identify the top-performing methods with respect to classification performance and identifying molecular drivers of disease. Furthermore, this could extend to regression tasks or clustering tasks. This project can greatly benefit the community, either by developing new methods to assess their strength or finding a method that best suits custom multiomics data.

-->


<!-- # One more thing -->

<!-- This will be Appendix B. -->
